---
title: "K-Means Clustering with Palmer Penguins"
author: "Namaah Hans"
date: today
format: html
---

## 1a. K-Means

We begin by loading the Palmer Penguins dataset and selecting only two numeric features: `bill_length_mm` and `flipper_length_mm`. These will be the input variables for clustering.

```{r}
library(tidyverse)
library(cluster)

# Load and clean data
penguins <- read_csv("palmer_penguins.csv") %>%
  drop_na(bill_length_mm, flipper_length_mm) %>%
  select(bill_length_mm, flipper_length_mm)

my_kmeans <- function(data, k, max_iter = 100){
  set.seed(123)
  n <- nrow(data)
  centers <- data[sample(1:n, k), ]
  cluster_assignments <- rep(0, n)

  for (i in 1:max_iter){
    # Assign clusters
    dists <- as.matrix(dist(rbind(centers, data)))[1:k, (k+1):(k+n)]
    cluster_assignments <- apply(dists, 2, which.min)

    # Recompute centers
    new_centers <- sapply(1:k, function(j){
      colMeans(data[cluster_assignments == j, ])
    })
    new_centers <- t(new_centers)

    if (all(centers == new_centers)) break
    centers <- new_centers
  }

  return(list(clusters = cluster_assignments, centers = centers))
}

result <- my_kmeans(penguins, k = 3)
penguins$cluster_custom <- as.factor(result$clusters)

ggplot(penguins, aes(x = bill_length_mm, y = flipper_length_mm, color = cluster_custom)) +
  geom_point(size = 2) +
  labs(title = "Custom K-Means Clustering (k = 3)")
```

The plot shows the clusters formed by my custom K-Means algorithm (k = 3). The separation looks reasonable, confirming that the implementation works as expected.

```{r}
k_builtin <- kmeans(penguins, centers = 3)
penguins$cluster_builtin <- as.factor(k_builtin$cluster)

ggplot(penguins, aes(x = bill_length_mm, y = flipper_length_mm, color = cluster_builtin)) +
  geom_point(size = 2) +
  labs(title = "Built-in KMeans Clustering (k = 3)")
```

The built-in KMeans function (k = 3) produces similar clusters, validating the results from the custom implementation.

```{r}
library(cluster)

wcss <- numeric()
silhouette_scores <- numeric()

for (k in 2:7) {
  set.seed(1)
  km_out <- kmeans(penguins, centers = k, nstart = 10)
  wcss[k] <- km_out$tot.withinss

  sil <- silhouette(km_out$cluster, dist(penguins))
  silhouette_scores[k] <- mean(sil[, 3])
}


# Elbow Plot
plot(2:7, wcss[2:7], type = "b", pch = 19,
     xlab = "Number of Clusters (k)",
     ylab = "Within-Cluster Sum of Squares",
     main = "Elbow Method")
```

The elbow plot suggests that 3 clusters is a good choice, as the rate of improvement in WCSS slows significantly after k = 3.

```{r}
# Silhouette Plot
plot(2:7, silhouette_scores[2:7], type = "b", pch = 19,
     xlab = "Number of Clusters (k)",
     ylab = "Silhouette Score",
     main = "Silhouette Analysis")

```

The silhouette scores are highest at k = 2, but k = 3 still provides a reasonable balance with better separation than higher k values.


## 2a. K Nearest Neighbors

```{r}
# gen data -----
set.seed(42)
n <- 100
x1 <- runif(n, -3, 3)
x2 <- runif(n, -3, 3)
x <- cbind(x1, x2)

# define a wiggly boundary
boundary <- sin(4*x1) + x1
y <- ifelse(x2 > boundary, 1, 0) |> as.factor()
dat <- data.frame(x1 = x1, x2 = x2, y = y)

### Plot the training data
```
We visualize the synthetic data with `x1` on the x-axis, `x2` on the y-axis, and color-coded by class `y`.

```{r}
library(ggplot2)

ggplot(dat, aes(x = x1, y = x2, color = y)) +
  geom_point(size = 2) +
  labs(title = "Training Data with Wiggly Boundary",
       x = "x1", y = "x2")
```

The plot shows two classes separated by a non-linear (wiggly) boundary, making it ideal for testing KNN performance.

```{r}
set.seed(99)
x1_test <- runif(n, -3, 3)
x2_test <- runif(n, -3, 3)
boundary_test <- sin(4*x1_test) + x1_test
y_test <- ifelse(x2_test > boundary_test, 1, 0) |> as.factor()
test_dat <- data.frame(x1 = x1_test, x2 = x2_test, y = y_test)



library(class)

k_values <- 1:30
accuracy <- numeric(length(k_values))

for (k in k_values) {
  pred <- knn(train = dat[, 1:2],
              test = test_dat[, 1:2],
              cl = dat$y,
              k = k)
  accuracy[k] <- mean(pred == test_dat$y)
}


plot(k_values, accuracy, type = "b", pch = 19,
     xlab = "k (Number of Neighbors)",
     ylab = "Test Set Accuracy",
     main = "KNN Accuracy for k = 1 to 30")

```
Test accuracy fluctuates across values of k, with the best performance observed around k = 29–30.